Survival Compass

### 2024-04-28 paper/paper.qmd

---
title: "Survival Compass"
subtitle: "Understanding the Impact of Pathologic Stage and Treatment Modalities on Lung Cancer Patients' Survival Post-Diagnosis."
author: 
  - Lexi Knight
thanks: "Code and data are available at: https://github.com/LexiKnight/Lung_Cancer/tree/main"
date: today
date-format: long 
abstract: "This study investigates the impact of pathologic stage and treatment modalities on lung cancer survival post-diagnosis. Analysis of patient data reveals significant correlation between the variables that make up pathologic stage, treatment type and survival outcomes. Notably, patients who demonstrate lower surivival rates are those with advanced spread, extensive lymph node involvement and tumors with extensive growth into nearby structures. Additionally pharmaceutical therapy is shown to be more effective in prolonging patient survial times than radiation therapy. These findings underscore the critical importance of early detection, tailored treatment strategies and ongoing research efforts to enhance lung cancer survival rates globally."
format:
  pdf:
    toc: true
number-sections: true
bibliography: references.bib
---

@@

Following tumor size, LN involvement and presence of distant metastasis categorization, the pathologic stage of the cancer is then determined, the most valuable prognostic facor in non-small cell lung cancer [@Eldridge]. Stages ranging from localized to distant spread dictate treatment strategies and prognosis, with curative options diminishing as the disease advances (see @sec-appendix for staging). Despite advancements in treatment modalities such as radiation therapy and chemotherapy, stage IV disease remains particularly challenging, emphasizing the importance of early detection and intervention [@Denton]. The earlier the cancer is found, that is the lower the pathologic stage, the greater the likelihood curative radiation therapy is an effective treatment [@Eldridge].

@@

### Data Cleaning
After obtaining the dataset, we selected the columns of interest namely; days to death post diagnosis, presence of distant metastasis, lymph node involvement, pathologic stage, tumor size and treatment type. Next, we renamed the columns, giving them meaningful names and excluded the data containing missing values. This left us with a sample of 382 lung cancer patients. Additionally, we converted the days to death column to numeric.Tests where included to ensure accuracy, reliability and validity of the datset for subsequent analysis and interpretation.

@@

@fig-one illustrates the survival curve for lung cancer patients post-diagnosis where frequency is on the y-axis and survival time in days is on the x-axis. There is a clear trend of a decrease in frequency of survival as time increases. Mortality is most abundant 500 days after diagnosis, that is equivalent to about a year and four months. After about 1500 days, just over four years, the trend line plateus.

@fig-one, the distribution of survival time post-diagnosis suggests that there is a high mortality rate shortly after diagnosis, thus most lung cancer patients die within one year of diagnosis (500 days). There is a steady decline in death and hence incline in survial up until four years (1500 days). The plateu after this indicates that there is a subgroup of patients who survive beyond four years. @sec-results explores which patients had longer survival.

@@

### pathologic Stage
Pathologic stage, a critical determinant of lung cancer prognosis, was classified according to the TNM (tumor, lymph node, metastasis) Cancer Staging. As depicted in @fig-two, illustrating the the distribution of pathologic stages with pathologic stage on the x-axis and percentage on the y-axis. The distribution of patients varied across different stages, with the highest proportion of patients diagnosed at StageIB. At this stage, the cancer is still found early however the tumor has regional spread to nearby lymph nodes. Interestingly, at diagnosis, patients had a fairly even spread across the different stages.

@@

```{r}
#| warning: false
#| echo: false
#| message: false
#| label: fig-three
#| fig-cap: The histogram shows the percentage of lung cancer patients by treamtent type.

@@

In this section, we aim to predict the survival outcomes of lung cancer patients post-diagnosis with a linear regression model framework. We consider several predictors including pathologic stage, lymph node involvement, presence of distant metastasis, tumor size, and treatment type.We specify the model and subsequently justify its appropriateness for our analysis.

@@

Linear regression models are most appropritate in predicting continuous outcomes. As survival time is continuous, this model allows us to quantify the relationships between these predictors and survival outcomes, providing valuable insights into the factors influencing the prognosis of lung cancer patients.

@@

Out variable of interest is survival time in lung cancer patient afte they have been diagnosed

@@

Stage III: cancer found in one lung and has spread to the chest and LN
  further further away from the lungs. Regional spread.

@@

M1b: metastasis to distant sites beyong lung and pleura, ex. brain, liver, bones. Distant.

>>>>

---
title: "Survival Compass"
subtitle: "Understanding the Impact of Pathologic Stage and Treatment Modalities on Lung Cancer Patients' Survival Post-Diagnosis."
author: 
  - Lexi Knight
thanks: "Code and data are available at: https://github.com/LexiKnight/Lung_Cancer/tree/main"
date: today
date-format: long 
abstract: "This study investigates the impact of pathologic stage and treatment modalities on lung cancer survival post-diagnosis. Analysis of patient data reveals significant correlation between the variables that make up pathologic stage, treatment type and survival outcomes. Notably, patients who demonstrate lower survival rates are those with advanced spread, extensive lymph node involvement and tumors with extensive growth into nearby structures. Additionally pharmaceutical therapy is shown to be more effective in prolonging patient survival times than radiation therapy. These findings underscore the critical importance of early detection, tailored treatment strategies and ongoing research efforts to enhance lung cancer survival rates globally."
format:
  pdf:
    toc: true
number-sections: true
bibliography: references.bib
---

@@

Following tumor size, LN involvement and presence of distant metastasis categorization, the pathologic stage of the cancer is then determined, the most valuable prognostic factor in non-small cell lung cancer [@Eldridge]. Stages ranging from localized to distant spread dictate treatment strategies and prognosis, with curative options diminishing as the disease advances (see @sec-appendix for staging). Despite advancements in treatment modalities such as radiation therapy and chemotherapy, stage IV disease remains particularly challenging, emphasizing the importance of early detection and intervention [@Denton]. The earlier the cancer is found, that is the lower the pathologic stage, the greater the likelihood curative radiation therapy is an effective treatment [@Eldridge].

@@

### Data Cleaning
After obtaining the dataset, we selected the columns of interest namely; days to death post diagnosis, presence of distant metastasis, lymph node involvement, pathologic stage, tumor size and treatment type. Next, we renamed the columns, giving them meaningful names and excluded the data containing missing values. This left us with a sample of 382 lung cancer patients. Additionally, we converted the days to death column to numeric. Tests where included to ensure accuracy, reliability and validity of the dataset for subsequent analysis and interpretation.

@@

@fig-one illustrates the survival curve for lung cancer patients post-diagnosis where frequency is on the y-axis and survival time in days is on the x-axis. There is a clear trend of a decrease in frequency of survival as time increases. Mortality is most abundant 500 days after diagnosis, that is equivalent to about a year and four months. After about 1500 days, just over four years, the trend line plateaus.

@fig-one, the distribution of survival time post-diagnosis suggests that there is a high mortality rate shortly after diagnosis, thus most lung cancer patients die within one year of diagnosis (500 days). There is a steady decline in death and hence incline in survival up until four years (1500 days). The plateau after this indicates that there is a subgroup of patients who survive beyond four years. @sec-results explores which patients had longer survival.

@@

### pathologic Stage
Pathologic stage, a critical determinant of lung cancer prognosis, was classified according to the TNM (tumor, lymph node, metastasis) Cancer Staging. As depicted in @fig-two, illustrating the distribution of pathologic stages with pathologic stage on the x-axis and percentage on the y-axis. The distribution of patients varied across different stages, with the highest proportion of patients diagnosed at Stage IB. At this stage, the cancer is still found early however the tumor has regional spread to nearby lymph nodes. Interestingly, at diagnosis, patients had a fairly even spread across the different stages.

@@

```{r}
#| warning: false
#| echo: false
#| message: false
#| label: fig-three
#| fig-cap: The histogram shows the percentage of lung cancer patients by treatment type.

@@

In this section, we aim to predict the survival outcomes of lung cancer patients post-diagnosis with a linear regression model framework. We consider several predictors including pathologic stage, lymph node involvement, presence of distant metastasis, tumor size, and treatment type. We specify the model and subsequently justify its appropriateness for our analysis.

@@

Linear regression models are most appropriate in predicting continuous outcomes. As survival time is continuous, this model allows us to quantify the relationships between these predictors and survival outcomes, providing valuable insights into the factors influencing the prognosis of lung cancer patients.

@@

Our variable of interest is survival time in lung cancer patient after they have been diagnosed

@@

Stage III: cancer found in one lung and has spread to the chest and LN
  further away from the lungs. Regional spread.

@@

M1b: metastasis to distant sites beyond lung and pleura, ex. brain, liver, bones. Distant.
